The genomic impacts of drift and selection for hybrid performance in maize

### 2015-08-27 main.tex
edits to Intro/Abstract

\abstract{
Although maize is naturally an outcrossing organism, modern breeding takes advantage of highly inbred lines that are used in controlled crosses to produce hybrids. 
The United States Department of Agriculture reciprocal recurrent selection experiment between the Iowa Stiff Stalk Synthetic (BSSS) and the Iowa Corn Borer Synthetic No. 1 (BSCB1) populations represents one of the longest \rev{running experiments to understand the genetic bases} for hybrid performance. 
To investigate the genomic impact of this selection program, \rev{we genotyped the progenitor lines and over 600 individuals across multiple cycles of selection using a genome-wide panel of $\sim 40,000$ SNPs.
We confirmed previous results showing a steady temporal decrease in genetic diversity within populations and a corresponding increase in differentiation between populations.  
Thanks to detailed historical information on experimental design, we were able to perform extensive simulations using founder haplotypes to replicate the experiment in the absence of selection.
These simulations demonstrate that while most of the observed reduction in genetic diversity can be attributed to genetic drift,   
heterozygosity in each population has fallen more than expected. 
We then took advantage of our high-density genotype data to identify extensive regions of haplotype fixation and trace haplotype ancestry to single founder inbred lines.
The vast majority of regions showing such evidence of selection differ between the two populations, providing evidence for the dominance model of heterosis.}
We discuss how this pattern is likely to occur during selection for hybrid performance, and how it poses challenges for dissecting the impacts of modern breeding and selection on the maize genome. }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% INTRO
\section*{Introduction} 
Hybrid maize, first developed in the early 20th century \citep{crow1998}, was so successful that within a few decades it had completely replaced mass-selected open pollinated varieties in the United States \citep{crabb1947hybrid}. 
The shift towards development of inbred lines based on their ability to generate good hybrids – referred to as ‘combining ability’ – constituted an abrupt change from the open-pollinated mass selection that breeders practiced for millennia \citep{anderson1944sources, troyer1999background}.
Maize inbred lines are now partitioned into separate ‘heterotic groups’ that maximize performance and hybrid vigor (heterosis) \rev{for yield} when inbreds from different heterotic groups are crossed with each other \citep{tracy2006historical}.

\rev{While the founders of these heterotic groups were not initially differentiated, m}ultiple studies with molecular markers have indicated that these heterotic groups have diverged genetically over time to become highly structured and isolated populations, resulting in a dramatic restructuring of population genetic variation \citep{duvick2004long, ho2005extent, feng2006temporal}. 
Advances in high throughput genotyping and the development of a maize reference genome now enable the observation of maize population structure at high marker density across the whole genome \citep{ganal2011a-large,chia2012maize}. 
These studies have examined a broad spectrum of germplasm at various points in the history of maize to search for the signals of population structure and artificial selection \citep{Hufford2012b, van2012historical,jiao2012genome}. 
Although selective sweeps remaining from domestication are clearly visible, the impact of selection during modern breeding appears comparatively small in terms of its impact on genomic diversity \citep{Hufford2012b,van2012historical}, despite steady, heritable improvement in phenotype \citep{duvick2005contribution}. 
The lack of distinct selection signals \rev{in broad germplasm collections may be due to population-specific selection within germplasm subgroups; in this case selection should be easier to detect in populations maintained in an individual program.} 
This \rev{possibliity} is supported by the measured success in identifying targets of selection in individual experimental populations \rev{under directional selection for specific phenotypes such as seed size and ear number} \citep{hirsch2014insights, beissinger2014genome}.

\rev{In this study, we examine a different experiment in which the  method and target of selection --- reciprocal recurrent selection for hybrid yield --- closely mirror those used in the generation of modern maize hybrids \citep{comstock1949breeding, duvick2004long}. 
Proposed by \citet{comstock1949breeding}, reciprocal recurrent selection simultaneously improves two populations for their ability to produce high yielding hybrids when individuals from the two populations are crossed. 
In this method, the two populations are isolated genetically and lines from each population are selected based on the phenotype of their cross-population hybrids. 
The superio lines within each population are then intermated to generate lines for the next cycle. 
In this way, the two populations simultaneously evolve improved combining ability with one another. 
The USDA-ARS at Ames, Iowa has conducted a reciprocal recurrent selection program (hereafter referred to as the Iowa RRS) with the Iowa Stiff Stalk Synthetic (BSSS) and Iowa Corn Borer Synthetic No. 1 (BSCB1) populations for 18 cycles of selection \citep{penny1971twenty}. 
This program represents one of the best-documented public experiments on selection for combining ability and hybrid performance. 
Extensive records on population sizes, breeding methods, selection differentials, and seed are available for all cycles of selection. 
This set of resources makes the Iowa RRS an ideal test case for the study of the genomic impact of hybrid breeding in maize.} 
The Iowa RRS experiment \rev{provides additional relevance} because lines derived from the BSSS population have had a major impact on the development of commercial hybrids \citep{duvick2004long, darrah1986}, the formation of modern heterotic groups \citep{troyer1999background, senior1998utility}, and the choice of a maize reference genome \citep{schnable2009the-b73-maize}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% INTRO

\section*{Methods}
\subsection*{The BSSS and BSCB1 recurrent selection program}
\rev{The Iowa RRS program was initiated in 1949 \citep{penny1971twenty}. 
The BSSS population was formed from 16 inbred lines selected for stalk strength in 1933 and 1934 (Sprague, 1946) and the BSCB1 population was formed from a set of 12 inbred lines \citep{hallauer1974registration}.} 
Founder lines (Table \ref{tab:founders}) from each population were randomly mated to create the `cycle 0' base population.
\rev{Testcross progenies were formed by self-pollinating 100 individuals in each population and simultaneously crossing the same 100 individuals onto 10 plants (as females) in the reciprocal population (e.g., a BSSS plant was self-pollinated and crossed to 10 females in BSCB1).  
Seed from the 10 females pollinated by a single male plant was bulked, and the 200 testcross progenies were grown in replicated multi-environment yield trials.  
Ten testcross families were selected from among the BSSS males (crossed to BSCB1 as a tester) and BSCB1 males (crossed to BSSS as a tester) based primarily on grain yield.  
Self-pollinated seed from the 10 selected males in each population were planted in the following season and plant-to-plant crosses made between each of the 45 possible pairs; seed from these crosses formed the cycle 1 populations.}

@@

\subsection*{Population structure and genetic diversity}
Founder inbreds and samples from cycles 0, 4, 8, 12, and 16 were genotyped at 39,261 SNPs that passed a set of quality filters and could be assigned collinear genetic and physical map positions (See Materials and Methods for details). 
Change in population structure throughout the Iowa RRS experiment can be observed visually by a principal component analysis (PCA). 
A joint analysis of individuals from all the selection cycles (Figure \ref{fig:pca}) clearly separates the BSSS and BSCB1 populations along the first axis of variation, with increasing separation as the experiment progressed. 
The second axis of variation primarily separates the cycles from one another within each population. 
There is no separation between the founders of the two populations, \rev{and projection of later cycles onto a PCA of the founders shows no distinction between BSSS and BSCB1 (Figure \ref{fig:sfounders}).
At cycle 0, however,} the BSSS population shows more divergence from the founders than does BSCB1, likely due to drift during either the population’s construction or subsequent maintenance. 
Structure continued to develop within each population over the course of the experiment. 
There is an especially wide gap between cycles 4 and 8, which correlates with the addition of an extra generation of self-pollination prior to selection at each cycle. 
The distance between cycles then decreases dramatically after cycle 8, and corresponding to the increased effective population size (See Materials and Methods).

@@

Our analysis of diversity in the Iowa RRS experiment reveals a steady loss of diversity in the BSSS and BSCB1 populations as they became increasingly differentiated from one another over time.
Principal component analysis shows that as the effective population size and the rates of inbreeding were altered, the rates of change in population structure were altered as well. 
These patterns of population structure, diversity, and differentiation between BSSS and BSCB1 can be largely reproduced by simulation without any selection, supporting the hypothesis that the majority of the genetic structure observed can be attributed to genetic drift alone, despite effective selection for phenotypic improvement. 
\rev{Similar observations have been recently made in other reciprocal recurrent selection programs, even in comparisons of multiple replicated populations \citep{romay2012effect,lamkey2014relative}.
These patterns differ markedly, however, from similar experimental evolution in  systems such as \emph{Drosophila} \citep{turner2011population, burke2010genome} and \emph{E. coli} \citep{tenaillon2012molecular} in which effective population sizes are maintained at levels much higher than in the present work and the effects of selection on diversity are readily discernible.}

>>>>

\abstract{
Although maize is naturally an outcrossing organism, modern breeding develops of highly inbred lines that are used in controlled crosses to produce hybrids. 
The United States Department of Agriculture reciprocal recurrent selection experiment between the Iowa Stiff Stalk Synthetic (BSSS) and the Iowa Corn Borer Synthetic No. 1 (BSCB1) populations represents one of the longest \rev{running experiments to understand the response to selection} \jpg{changed the wording here, because the experiment's goal was more broad than undesrtanding the genetic basis} for hybrid performance. 
To investigate the genomic impact of this selection program, \rev{we genotyped the progenitor lines and over 600 individuals across multiple cycles of selection using a genome-wide panel of $\sim 40,000$ SNPs.
We confirmed previous results showing a steady temporal decrease in genetic diversity within populations and a corresponding increase in differentiation between populations.  
Thanks to detailed historical information on experimental design, we were able to perform extensive simulations using founder haplotypes to replicate the experiment in the absence of selection.
These simulations demonstrate that while most of the observed reduction in genetic diversity can be attributed to genetic drift,   
heterozygosity in each population has fallen more than expected. 
We then took advantage of our high-density genotype data to identify extensive regions of haplotype fixation and trace haplotype ancestry to single founder inbred lines.
The vast majority of regions showing such evidence of selection differ between the two populations, providing evidence for the dominance model of heterosis.}
We discuss how this pattern is likely to occur during selection for hybrid performance, and how it poses challenges for dissecting the impacts of modern breeding and selection on the maize genome. }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% INTRO
\section*{Introduction} 
Hybrid maize, first developed in the early 20th century \citep{crow1998}, rapidly and completely replaced mass-selected open pollinated varieties in the United States \citep{crabb1947hybrid}. 
The shift towards development of inbred lines based on their ability to generate good hybrids –-- referred to as ‘combining ability’ –-- constituted an abrupt change from the open-pollinated mass selection that breeders practiced for millennia \citep{anderson1944sources, troyer1999background}.
Maize inbred lines are now partitioned into separate ‘heterotic groups’ that maximize performance and hybrid vigor (heterosis) \rev{for yield} when inbreds from different heterotic groups are crossed with each other \citep{tracy2006historical}.

\rev{While the founders of these heterotic groups were not initially differentiated, m}ultiple studies with molecular markers have indicated that these heterotic groups have diverged genetically over time to become highly structured and isolated populations, resulting in a dramatic restructuring of population genetic variation \citep{duvick2004long, ho2005extent, feng2006temporal}. 
Advances in high throughput genotyping and the development of a maize reference genome now enable the observation of maize population structure at high marker density across the whole genome \citep{ganal2011a-large,chia2012maize}. 
These studies have examined a broad spectrum of germplasm at various points in the history of maize to search for the signals of population structure and artificial selection \citep{Hufford2012b, van2012historical,jiao2012genome}. 
Although selective sweeps from the initial domestication of maize are clearly visible, localized genomic signals of selection during modern breeding remain difficult to observe \citep{Hufford2012b,van2012historical} despite steady, heritable improvement in phenotype \citep{duvick2005contribution}. 
The lack of distinct selection signals \rev{in broad germplasm collections may be due to population-specific selection within germplasm subgroups; in this case selection should be easier to detect in populations maintained in an individual program.} 
This \rev{possibliity} is supported by the measured success in identifying targets of selection in individual experimental populations \rev{under directional selection for specific phenotypes such as seed size and ear number} \citep{hirsch2014insights, beissinger2014genome}.

\rev{In this study, we examine a different experiment in which the method and target of selection --- reciprocal recurrent selection for hybrid yield --- closely mirror those used in the generation of modern maize hybrids \citep{comstock1949breeding, duvick2004long}. 
Reciprocal recurrent selection is a method, initially proposed by \citet{comstock1949breeding}, in which lines from two populations are evaluated based on the phenotype of the hybrids each line produces when crossed with the opposing population.
The selected lines are then intermated within each population to generate lines for the next cycle of recurrent selection.
%Reciprocal recurrent selection simultaneously improves two populations for their ability to produce high yielding hybrids when individuals from the two populations are crossed. 
This procedure results in two closed and genetically isolated populations that simultaneously evolve improved combining ability with one another. 
The USDA-ARS at Ames, Iowa has conducted a reciprocal recurrent selection program (hereafter referred to as the Iowa RRS) with the Iowa Stiff Stalk Synthetic (BSSS) and Iowa Corn Borer Synthetic No. 1 (BSCB1) populations for 18 cycles of selection for hybrid yield \citep{penny1971twenty}. \jpg{I think we need Jode's reference here as well if we are going to mention it has gone through 18 cycles.}
This program represents one of the best-documented public experiments on selection for combining ability and hybrid performance. 
Extensive records on population sizes, breeding methods, selection differentials, and seed are available for all cycles of selection. 
This set of resources makes the Iowa RRS an ideal test case for the study of the genomic impact of hybrid breeding in maize.} 
The Iowa RRS experiment \rev{provides additional relevance} because lines derived from the BSSS population have had a major impact on the development of commercial hybrids \citep{duvick2004long, darrah1986}, the formation of modern heterotic groups \citep{troyer1999background, senior1998utility}, and the choice of a maize reference genome \citep{schnable2009the-b73-maize}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% INTRO

\section*{Methods}
\subsection*{The BSSS and BSCB1 recurrent selection program}
\rev{The Iowa RRS program was initiated in 1949 \citep{penny1971twenty}. 
The BSSS population was formed from 16 inbred lines selected for stalk strength in 1933 and 1934 (Sprague, 1946) and the BSCB1 population was formed from a set of 12 inbred lines \citep{hallauer1974registration}.} 
The founder inbreds (Table \ref{tab:founders}) were randomly mated to create each `cycle 0' base population.
\rev{Testcross progenies were formed by self-pollinating 100 individuals in each population and simultaneously crossing the same 100 individuals onto 10 plants (as females) in the reciprocal population (e.g., a BSSS plant was self-pollinated and crossed to 10 females in BSCB1).  
Seed from the 10 females pollinated by a single male plant was bulked, and the 200 testcross progenies were grown in replicated multi-environment yield trials.  
Ten testcross families were selected from among the BSSS males (crossed to BSCB1 as a tester) and BSCB1 males (crossed to BSSS as a tester) based primarily on grain yield.  
Self-pollinated seed from the 10 selected males in each population were planted in the following season and plant-to-plant crosses made between each of the 45 possible pairs; seed from these crosses formed the cycle 1 populations.}

@@

\subsection*{Population structure and genetic diversity}
Founder inbreds and samples from cycles 0, 4, 8, 12, and 16 were genotyped at 39,261 SNPs that passed a set of quality filters and could be assigned collinear genetic and physical map positions (See Materials and Methods for details). 
Change in population structure throughout the Iowa RRS experiment can be observed visually by a principal component analysis (PCA). 
Analysis of individuals from all the selection cycles (Figure \ref{fig:pca}) clearly separates the BSSS and BSCB1 populations along the first axis of variation, with increasing separation as the experiment progressed. 
The second axis of variation primarily separates the cycles from one another within each population. 
There is no separation between the founders of the two populations, \rev{and projection of later cycles onto a PCA of the founders shows no distinction between BSSS and BSCB1 (Figure \ref{fig:sfounders}).
At cycle 0, however,} the BSSS population shows more divergence from the founders than does BSCB1, likely due to drift during either the population’s construction or subsequent maintenance. 
Structure continued to develop within each population over the course of the experiment. 
There is an especially wide gap between cycles 4 and 8, which correlates with the addition of an extra generation of self-pollination prior to selection at each cycle. 
The distance between cycles then decreases dramatically after cycle 8, and corresponding to the increased effective population size (See Materials and Methods).

@@

Our analysis of the Iowa RRS experiment reveals a steady loss of diversity in the BSSS and BSCB1 populations as they became increasingly differentiated from one another over time.
Principal component analysis shows that as the effective population size and the rates of inbreeding were altered, the rates of change in population structure were altered as well. 
These patterns of population structure, diversity, and differentiation between BSSS and BSCB1 can be largely reproduced by simulation without any selection, supporting the hypothesis that the majority of the genetic structure observed can be attributed to genetic drift alone, despite effective selection for phenotypic improvement. 
\rev{Similar observations have been recently made in other reciprocal recurrent selection programs, even in comparisons of multiple replicated populations \citep{romay2012effect,lamkey2014relative}.
These patterns differ markedly, however, from similar experimental evolution in  systems such as \emph{Drosophila} \citep{turner2011population, burke2010genome} and \emph{E. coli} \citep{tenaillon2012molecular} in which effective population sizes are maintained at levels much higher than in the present work and the effects of selection on diversity are readily discernible.}
